Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2020-12-04 analysis/paper/paper.Rmd
Commit for push!
split the codes for each plot
upgraded R studio and tested it

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human]. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- generally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) --> 
<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths. In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes. The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance]. The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

<!-- the last sentense is a bit long-->
<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->

The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca. 36,000 cal BP) [@yamaoka2012use; @izuho201421]. The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421]. These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421].

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that --> 
<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].

The origin of the stemmed point is one of the most frequently discussed topics among Paleolithic research in Korea. Especially, this debate includes questions about the origin of the Late Paleolithic period in Korea, the existence and nature of Middle Paleolithic technology, and reconstructions of anatomically modern human dispersals in eastern Asia. Most Korean archaeologists had assumed that the new blade technology represented came from Shuidonggou in the northeast China because this site has one of the earliest blade industries (around 35,000 BP) in northeast Asia and it is geographically close to the Korean peninsula [seong2008tanged; @lee2013current]. However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence]. The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and ‘heterogenic’ migration [@bae2010origin]. The ‘in situ’ model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence. Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim. In his view, the blade industry represents new technology while flakes indicate a continuing local one. He argues that the increased ratio of blades in lithic assemblages shows an expansion of the new technology. The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins. While Seong’s model (2009) proposes that the ‘heterogenic’ character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China. Bae’s model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early]. 
<!--- I don't know how to connect this debate to our research. It is really needed? --->

@@

To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type. The elevation is the main proxy for environmental information in this research. The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [####Han 2008####; @Chang_2013]. Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ [###Han 2008###; ####Seong 2008####; ###Choi 2011####; @Chang_2013; @im2015study; @bak2017late]. In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula. In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017]. In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk. This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.

@@

Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].

@@

In terms of the composition of toolkits, we count kind of tools as well as the amount of debitabitage to search for the change in the toolkits around the transition period. The tools represent paleo hunter-gatherers’ adaptive strategies. We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization [@brantingham2003neutral].

Umap: <- depending on whether we are gonna add umap (or PCA) analysis result

@@

# comparable figure to the raw material one (BM created)

@@

```
Artifact density and retouch <- maybe we need to add the result of retouch, separately 
We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers. The result shows that there is no clear pattern during the transition period and then it became more expedient at the end of the transition (Fig 5. The combination of artifact density and retouch). That is to say, the hunter-gatherer became either more residential or the majority of the site that this research studies were used as base camps or residence rather than shortly occupied camps for certain tasks.

@@

```
Umap

@@

The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. Therefore, we can assume that the technology would play an important role to survive in more risky environments.

@@

To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time]. There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated. It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp. Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.

@@

Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in  https://www.jstor.org/stable/4128419

@@

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

>>>>

The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new; @seong2008tanged; @chang2013human]. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- generally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) -->

<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged; @bae2010origin; @Bae_2010; @bae2017origin].

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths. In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes. The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance]. The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

<!-- the last sentense is a bit long-->

<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->

The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca. 36,000 cal BP) [@yamaoka2012use; @izuho201421]. The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliets sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421].

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that -->

\<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; \#\#\#\# Park 2012 \#\#\#\#; @seong2015diversity; @bae2017origin].

The origin of the stemmed point is one of the most frequently discussed topics among Paleolithic research in Korea. Especially, this debate includes questions about the origin of the Late Paleolithic period in Korea, the existence and nature of Middle Paleolithic technology, and reconstructions of anatomically modern human dispersals in eastern Asia. Most Korean archaeologists had assumed that the new blade technology represented came from Shuidonggou in the northeast China because this site has one of the earliest blade industries (around 35,000 BP) in northeast Asia and it is geographically close to the Korean peninsula [seong2008tanged; @lee2013current]. However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence]. The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and 'heterogenic' migration [@bae2010origin]. The 'in situ' model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence. Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim. In his view, the blade industry represents new technology while flakes indicate a continuing local one. He argues that the increased ratio of blades in lithic assemblages shows an expansion of the new technology. The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins. While Seong's model (2009) proposes that the 'heterogenic' character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China. Bae's model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early]. <!--- I don't know how to connect this debate to our research. It is really needed? --->

@@

To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type. The elevation is the main proxy for environmental information in this research. The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [\#\#\#\#Han 2008\#\#\#\#; @Chang_2013]. Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as 'overall cooler and drier during MIS 3' [\#\#\#Han 2008\#\#\#; \#\#\#\#Seong 2008\#\#\#\#; \#\#\#Choi 2011\#\#\#\#; @Chang_2013; @im2015study; @bak2017late]. In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula. In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017]. In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk. This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.

@@

Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents "base camps" or "residences", while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].

@@

In terms of the composition of toolkits, we count kind of tools as well as the amount of debitabitage to search for the change in the toolkits around the transition period. The tools represent paleo hunter-gatherers' adaptive strategies. We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization [@brantingham2003neutral].

Umap: \<- depending on whether we are gonna add umap (or PCA) analysis result

@@

# comparable figure to the raw material one

@@

Artifact density and retouch \<- maybe we need to add the result of retouch, separately We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers. The result shows that there is no clear pattern during the transition period and then it became more expedient at the end of the transition (Fig 5. The combination of artifact density and retouch). That is to say, the hunter-gatherer became either more residential or the majority of the site that this research studies were used as base camps or residence rather than shortly occupied camps for certain tasks.

@@

Umap

@@

The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers' mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. Therefore, we can assume that the technology would play an important role to survive in more risky environments.

@@

To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching; @binford1979organization; @torrence1989time]. There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated. It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp. Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.

@@

Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in <https://www.jstor.org/stable/4128419>

@@

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::
